---
title: "Bioreactor time-course analysis"
author: "Olivier Chapleur"
output:
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 6
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(fig.align = "center")
```

```{r ,echo=FALSE,warning=FALSE, message=FALSE}
rm(list=ls(all=TRUE))
#loading of the packages
library(lmms)
library(mixOmics)
library(knitr)
library(reshape2)
library(tidyverse)


```

# Data

In this data set, **three bioreactors** with similar performances were considered as replicates. Different parameters were measured accross time in the three bioreactors.

> Performance data: 
Based on chemical measurement, the time course evolution of a set of parameters was measured (CH4, C02, acetate, propionate). 

> Metabolites data: 
The time course evolution of 20 selected metabolites was measured with GCMS. 


> Microbial data: 
DNA from samples taken across time was extracted and sequenced. (16S metabarcoding). 



```{r, echo=FALSE}
#loading of the data
load('../Data/bioreactor.RData')
source('../Rscripts/spls.results.R')
source("../Rscripts/filter.R")
source("../Rscripts/block.spls.results.R")

# performance data: acetate  CH4  C02  propionate
# 16S data: abundance (raw counts), OTU_metadata (metadata associated to the OTUs), taxo (OTUs taxonomy)
# metabolomic data: GCMS (intensity of the metabolites), metadata_GCMS (metadata associated to the metabolites)
```

# Data preprocessing



>Metabolites (GCMS) data are log transformed. 

>Microbial data 

1) are filtered (only OTUs with at least 1% of abundance in at least 1 sample are kept = 51 OTUs).

2) a count of 1 sequence is added to each sample/OTU (to avoid 0 in the datamatrix)  

3) relative abundance is calculated 

4) obtain data is clr transformed

>Performance data is not transformed.


```{r, echo=FALSE, eval=TRUE}
#GCMS data are log transformed
GCMS.log=log(GCMS+1,10)
```

```{r, echo=FALSE, eval=TRUE}
# function for TSS transformation of the data
TSS.divide = function(x){
  x/sum(x)
}
# function is applied to each row to calculate the percentage of each OTU in the samples
pc.abundance = (apply(abundance, 2, TSS.divide))


#Only OTUs present at at least percent% in one of the samples are kept.
percent=0.01 # threshold for filtering (1%)
pc.abundance.sel = apply(subset((abundance),apply(pc.abundance,1, function(x) max(x,na.rm=TRUE))>percent)+1,2,TSS.divide)

#data are clr transformed
clr.abundance=logratio.transfo(t(pc.abundance.sel), logratio = 'CLR', offset = 0)



cat("There are", nrow(pc.abundance.sel), "OTUs after", percent, "% filter")

```

# Spline smoothing

All the data are modelled with spline smoothing with the Linear Mixed Model Splines framework (package `lmms`).

As a reminder, the LMMS modeling step tests 4 different models for each OTUs. 0 = linear model, 1 = linear
mixed effect model spline (LMMS) with defined basis, 2 = LMMS taking subject-specific random intercept, 3
= LMMS with subject specific intercept and slope.



## Metabolites

```{r, echo=FALSE, message=FALSE,eval=TRUE, warning=FALSE}
#spline smoothing of GCMS data
lmms.metabolites <- lmmSpline(data=(GCMS.log),time=metadata_GCMS$Number_of_days,sampleID=metadata_GCMS$rep_reactor,keepModels=T,timePredict=c(10:57),basis="p-spline") # values are predicted for every date
data.metabolites =lmms.metabolites@predSpline
data.metabolites=t(data.metabolites[,colnames(data.metabolites)%in%c(10,13,16,19,22,25,28,31,34,37,40,43,46,50,53,57)]) # only a few dates are kept (same dates for all the data)


summary(lmms.metabolites)# to get a summary of the lmms analysis
```

For the metabolites data, 10 molecules were modelled with a straight line, 4 with LMMS with subject-specific random intercept, 6 with LMMS with subject specific intercept and slope.

## OTUs

```{r, echo=FALSE, message=FALSE,eval=TRUE, warning=FALSE}
#spline for the OTUs
lmms.OTUs <- lmmSpline(data=(clr.abundance),time=OTU_metadata$Number_of_days,sampleID=OTU_metadata$reactor_id,keepModels=T,timePredict=c(10:57),basis="p-spline") # values are predicted for every date
data.OTUS=lmms.OTUs@predSpline
data.OTUS=t(data.OTUS[,colnames(data.OTUS)%in%c(10,13,16,19,22,25,28,31,34,37,40,43,46,50,53,57)]) # only a few dates are kept (same dates for all the data)

summary(lmms.OTUs)# to get a summary of the lmms analysis
```

For the microbiome data, 30 OTUs were modelled with a straight line, 19 with LMMS and 2 with LMMS with subject specific random intercept.

## Performance data

```{r, echo=FALSE, message=FALSE,eval=TRUE, warning=FALSE}
# we have a different number of time points and replicates for the performance data. 

#we first reshape the data to create one table for each performance data
melt.acetate=melt(as.matrix(acetate))
melt.propionate=melt(as.matrix(propionate))
melt.CH4=melt(as.matrix(CH4))
melt.CO2=melt(as.matrix(CO2))


#one table for liquid compound concentration (acetate and propionate), same number of time points
melt.liq=data.frame('time'=melt.acetate$Var1, 'sampleID'=melt.acetate$Var2,'acetate'=melt.acetate$value, 'propionate'=melt.propionate$value)

#one table for gas production (CH4 and CO2), same number of time points
melt.perf=data.frame('time'=melt.CH4$Var1, 'sampleID'=melt.CH4$Var2,'CH4'=melt.CH4$value, 'CO2'=melt.CO2$value)


#spline smoothing of this data 
lmms.liq <- lmmSpline(data=cbind(melt.liq$acetate, melt.liq$propionate),time=melt.liq$time,sampleID=melt.liq$sampleID,keepModels=T,timePredict=c(10:57),basis="p-spline")

lmms.perf <- lmmSpline(data=cbind(melt.perf$CH4, melt.perf$CO2),time=melt.perf$time,sampleID=melt.perf$sampleID,keepModels=T,timePredict=c(10:57),basis="p-spline")

#prediction after spline smoothing
liq.data=t(lmms.liq@predSpline[,colnames(lmms.liq@predSpline)%in%c(10,13,16,19,22,25,28,31,34,37,40,43,46,50,53,57)]) 
perf.data=t(lmms.perf@predSpline[,colnames(lmms.perf@predSpline)%in%c(10,13,16,19,22,25,28,31,34,37,40,43,46,50,53,57)])
colnames(liq.data)=c("acetate",'propionate')
colnames(perf.data)=c('CH4','CO2')



summary(lmms.liq)# to get a summary of the lmms analysis
summary(lmms.perf)# to get a summary of the lmms analysis
```

Acetate and Propionate were modelled with LMMS with subject specific random intercept. CH4 and CO2 were modelled with LMMS with subject specific intercept and slope.

# Filtering of the obtained profiles

Straight line modelling can occur when the inter-individual variation is too high. To remove the noisy profiles,
we first use the Breusch-Pagan test, which tests the homo-sedasticity of the residues. We then add a filter on
the mean squared error to reduce the dispersion of the residues around the line.

## OTUs

The 44 selected OTUs are listed below as well as the results of the filtering test.

```{r, echo=FALSE, eval=TRUE}

class(clr.abundance) <- c('matrix')

# to use the filtering function, sample IDs should be named as follows SAMPLE_TIME
row.names(clr.abundance)=paste(OTU_metadata$initial_CONCENTRATION,'_',  OTU_metadata$Number_of_days, sep="")


clr.abundance=data.frame(clr.abundance) #data table must be a data frame with molecule in columns


OTU.filter=wrapper.filter.splines(clr.abundance,lmms.OTUs,stationnarity.test = FALSE, homoskedasticity = TRUE,MSE.filter = TRUE)
#get_MSE(clr.abundance,lmms.OTUs )

# ls(OTU.filter) # what is returned by filtering function 
OTU.filter$to_keep ## OTUs to keep
#kable(OTU.filter$res.filter) ## for each molecule, indicate if the different test were passed, and with which model the molecule has been modelled


summary(OTU.filter$res.filter[,c(2,4)]) ## summary of the test


data.OTUS.kept=data.OTUS[,colnames(data.OTUS)%in%as.matrix(OTU.filter$to_keep)] # kept OTUs after filtering

matplot(x=matrix(rep(rownames(data.OTUS.kept), length(as.matrix(OTU.filter$to_keep))), ncol=length(as.matrix(OTU.filter$to_keep))), y=scale(data.OTUS.kept), type = 'l', main='plot of scaled kept OTUs', xlab='time') # plot of the kept OTUs

rejected.OTU.name=colnames(data.OTUS)[!colnames(data.OTUS)%in%as.matrix(OTU.filter$to_keep)] ### OTUs rejected
data.OTUS.rejected=data.OTUS[,!colnames(data.OTUS)%in%as.matrix(OTU.filter$to_keep)]

#matplot(x=matrix(rep(rownames(data.OTUS.rejected), length(as.matrix(rejected.OTU.name))), ncol=length(as.matrix(rejected.OTU.name))), y=scale(data.OTUS.rejected), type = 'l', main='plot of scaled rejected OTUs', xlab='time') # plot of the rejected OTUs

```

The filtering step removed 7 OTUs. 
In the above graph, time profiles of the selected OTUs are displayed.

## Metabolites

The 16 selected metabolites are listed below as well as the results of the filtering test.
```{r, echo=FALSE, eval=TRUE}

row.names(GCMS.log)=paste(metadata_GCMS$rep_reactor,'_',  metadata_GCMS$Number_of_days, sep="")# to use the filtering function, sample IDs should be named as follows SAMPLE_TIME



metabolites.filter=wrapper.filter.splines(GCMS.log,lmms.metabolites,stationnarity.test = FALSE, homoskedasticity = TRUE,MSE.filter = FALSE)

metabolites.filter$to_keep #molecules to keep
#metabolites.filter$res.filter
summary(metabolites.filter$res.filter)


kept.metabolites.name=metabolites.filter$res.filter[metabolites.filter$res.filter$BP.test,1] ### metabolites to keep
data.metabolites.initial=data.metabolites

data.metabolites.kept=data.metabolites[,colnames(data.metabolites)%in%as.matrix(metabolites.filter$to_keep)]


matplot(x=matrix(rep(rownames(data.metabolites.kept), length((metabolites.filter$to_keep))), ncol=length((metabolites.filter$to_keep))), y=scale(data.metabolites.kept), type = 'l', main='plot of scaled kept metabolites', xlab='time')

rejected.metabolites.name=metabolites.filter$res.filter[!colnames(data.metabolites)%in%as.matrix(metabolites.filter$to_keep),1]  ### molecules rejected

data.metabolites.rejected=data.metabolites[,!colnames(data.metabolites)%in%as.matrix(metabolites.filter$to_keep)] # profiles of the rejected molecules
 

#matplot(x=matrix(rep(rownames(data.metabolites.rejected), length(as.matrix(rejected.metabolites.name))), ncol=length(as.matrix(rejected.metabolites.name))), y=scale(data.metabolites.rejected), type = 'l', main='plot of scaled rejected metabolites', xlab='time') #plot of the rejected profiles

```

The filtering steps removed 4 metabolites. 
In the above graph, time profiles of the selected metabolites are displayed.


# sPLS

In the following section, we use *Projection on Latent Structures (PLS)* to cluster both OTU and metabolite time profiles.
We also use *sparse PLS* to identify a biological signature per cluster.

In the graphs below, time points are represented as points placed according to their projection in the
smaller subspace spanned by the components of the sPLS. They allow to visualize the similarities (the points
are grouped together) and the dissimilarities between the times.


```{r, echo=FALSE, results='asis',fig.height=4, eval=TRUE}

 
 X1 =data.OTUS.kept
 X2=data.metabolites.kept
 
  ncomp = 2
 

spls.res = spls(X = X1, Y = X2, keepX=c(20,12), keepY=c(8,8), ncomp = ncomp, mode = 'regression', scale = TRUE) # keepX keepY values can be modified
```
```{r,  echo=FALSE, results='asis',fig.height=4, eval=TRUE}
plotIndiv(spls.res) # individual plot
```

We also use an *Arrow plot* to represent the similarity between the 2 datasets.
Each arrow corresponds to one time. The start of the arrow indicates the location of the time in X (OTUs) in one plot, and the tip the location of the same time in Y (metabolites) in the other plot. 
Short arrows indicate if both data sets strongly agree and long arrows a disagreement between the two data sets.

```{r, echo=FALSE, results='asis',fig.height=4, eval=TRUE}
plotArrow(spls.res) # i
```

The contribution of each OTU in the construction of the new components can be displayed on the circle of
correlations plot. On this graph, the strongly correlated OTUs are projected in the same direction. We use
this information to build trajectory clusters.

```{r, echo=FALSE, results='asis',fig.height=4, eval=TRUE}
plotVar(spls.res) #variable plot
```



```{r, echo=FALSE}

#profiles of the variables selected on each axis for each block are extracted. (spls.result function)

summary.X1= spls.result(data = scale(X1), block = 'X', spls.res)
summary.X2 = spls.result(data = scale(X2), block = 'Y', spls.res)

#data are scaled in the function
```

In the graph bellow, we represent the mean profiles for each block.

```{r, echo=FALSE, results='asis',fig.height=4, eval=TRUE, warning=F}

# representation of the mean profiles for each block (means calculated per sign and based on original but scaled data provided in spls.result)
meltm = melt(t(cbind(summary.X1$spls.means, summary.X2$spls.means)))
meltm$comp = substr(meltm$Var1, 1, 5) 
meltm$data = rep(c(rep('OTU', ncol(summary.X1$spls.means)), rep('Metabolites', ncol(summary.X2$spls.means))), nrow(data.OTUS))
print(ggplot(data=meltm, aes(x=Var2, y=value, group = data, color = data)) + geom_line() + facet_wrap( ~ Var1) + ggtitle('sPLS clusters, mean profiles') + xlab('time') +ylab('scaled value'))
```

We have 4 clusters with sPLS. The time profiles for both OTUs and metabolites are represented by cluster.
The clusters are separated according to component and contibution. On the first component (first column), OTUs and metabolites with a negative contribution on the component 1 are displayed on the first row.
OTUs and metabolites with a positive contribution on the component 1 are displayed on the second row.
The same applies to OTUs and metébolites selected on the second component.


```{r, echo=FALSE, results='asis',fig.height=4, eval=TRUE, warning=F}

# all profiles can be represented
# --------------------
# melt data for ggplot
melt.data = melt(t(cbind(scale(X1[, names(summary.X1$var.selected)]), scale(X2[, names(summary.X2$var.selected)]))))
#melt.data = melt(t(cbind((X1[, names(summary.X1$var.selected)]), (X2[, names(summary.X2$var.selected)]))))
melt.data$comp= as.factor(rep(c(summary.X1$var.selected, summary.X2$var.selected), nrow(data.OTUS)))  # filling up the melted values by repeating comp and sign
melt.data$sign= as.factor(rep(c(summary.X1$var.sign, summary.X2$var.sign), nrow(data.OTUS)))
melt.data$data = rep(c(rep('OTUs', length(summary.X1$var.selected)), rep('Metabolites', length(summary.X2$var.selected))), nrow(data.OTUS))



color=c("tomato","slateblue")
names(color)=c("OTUs","Metabolites")

#png("spls_bioreactor.png", width = 3000, height = 1800, res = 400,pointsize=12)

print(ggplot(melt.data, aes(x=Var2, y=value, group = interaction(Var1,data), color = data)) +scale_color_manual(values=color)+ geom_line(lwd=0.2) +   facet_grid(sign~comp, labeller=labeller(sign = c(pos="Positive",neg="Negative"), comp=c('1'="Component 1", '2'="Component 2", '3'='Component 3')))+ ggtitle('                       sPLS, correlated data across time') + xlab('Number of days')+ scale_y_continuous(limits = c(-2, 2)) +ylab('Scaled value'))

#dev.off()

# print(ggplot(data=melt.data, aes(x=Var2, y=value, group = Var1, color = data)) + geom_line() + facet_wrap( ~ comp) + ggtitle('sPLS') + xlab('time'))
# # can change color = comp to color = Var1 to see the name of the vaiables selected
# 
# # subset only on a particular component and represent pos / neg shapes
# print(ggplot(subset(melt.data, comp == 1), aes(x=Var2, y=value, group = Var1, color = data)) + geom_line() + facet_wrap( ~ sign) + ggtitle('sPLS comp1') + xlab('time'))
# 
# 
# # subset only on a particular component and represent pos / neg shapes
# print(ggplot(subset(melt.data, comp == 2), aes(x=Var2, y=value, group = Var1, color = data)) + geom_line() + facet_wrap( ~ sign) + ggtitle('sPLS comp2') + xlab('time'))


##??? who's who 
## name of the different OTUs is extracted, as well as the related taxonomic information 
signcomp=paste(summary.X1$var.selected, summary.X1$var.sign) 
names(signcomp)=names(summary.X1$var.sign)
names.1neg=names(signcomp)[(signcomp=="1 neg")]
table.1neg=(taxo[names.1neg,])
names.1pos=names(signcomp)[(signcomp=="1 pos")]
table.1pos=(taxo[names.1pos,])
names.2neg=names(signcomp)[(signcomp=="2 neg")]
table.2neg=(taxo[names.2neg,])
names.2pos=names(signcomp)[(signcomp=="2 pos")]
table.2pos=(taxo[names.2pos,])

# same for metabolomic data 

signcompX2=paste(summary.X2$var.selected, summary.X2$var.sign) 
names(signcompX2)=names(summary.X2$var.sign)
names.1negX2=names(signcompX2)[(signcompX2=="1 neg")]
names.1posX2=names(signcompX2)[(signcompX2=="1 pos")]
names.2negX2=names(signcompX2)[(signcompX2=="2 neg")]
names.2posX2=names(signcompX2)[(signcompX2=="2 pos")]
```

## Proportionality analysis

Interpretation based on correlations between profiles must be made with caution as it is highly likely to be
spurious. Proportional distances has been proposed as an alternative to measure association.

In the following graphs, we represent all the proportionality distance $\varphi_s$ within clusters and the distance of features inside the clusters with entire background set. We first splited the analysis by omics data type and we computed the distance with the merged data after.

```{r, echo=F, warning=F, message=F}
library(propr)
load("../Data/bioreactor_clusters.RData")
source("../Rscripts/unscale.R")

cluster.bioR <- cluster.pls
OTU.bioR <- spls.res$X %>% unscale %>% `+`(abs(min(.)))
phs.OTU <- propr::propr(OTU.bioR, metric = 'phs')@matrix

metabo.bioR <- spls.res$Y %>% unscale %>% `+`(abs(min(.)))
phs.metabo <- propr::propr(metabo.bioR, metric = 'phs')@matrix

metabo.bioR <- spls.res$Y %>% unscale %>% `+`(abs(min(.)))
phs.metabo <- propr::propr(metabo.bioR, metric = 'phs')@matrix

res.phs.OTU <- phs.OTU %>% as.data.frame %>% rownames_to_column("feature1") %>%
    gather(feature2, value, -feature1) %>%
    filter(feature1 %in% cluster.bioR$molecule) %>%
    filter(feature2 %in% cluster.bioR$molecule) %>%
    left_join(cluster.bioR %>% filter(block == "OTU") %>% dplyr::select(-block), 
              by = c("feature1" = "molecule")) %>%
    rename(cluster1 = cluster) %>%
    left_join(cluster.bioR %>% filter(block == "OTU"), 
              by = c("feature2" = "molecule")) %>%
    rename(cluster2 = cluster) %>%
    mutate(insideout = ifelse(cluster1 == cluster2, "inside", "outside"))

res.phs.metabo <- phs.metabo %>% as.data.frame %>% rownames_to_column("feature1") %>%
    gather(feature2, value, -feature1) %>%
    filter(feature1 %in% cluster.bioR$molecule) %>%
    filter(feature2 %in% cluster.bioR$molecule) %>%
    left_join(cluster.bioR %>% filter(block == "Metabolites") %>% dplyr::select(-block), 
              by = c("feature1" = "molecule")) %>%
    rename(cluster1 = cluster) %>%
    left_join(cluster.bioR %>% filter(block == "Metabolites"), 
              by = c("feature2" = "molecule")) %>%
    rename(cluster2 = cluster) %>%
    mutate(insideout = ifelse(cluster1 == cluster2, "inside", "outside"))

res.rbind <- rbind(res.phs.OTU, res.phs.metabo) %>%
    mutate(insideout = ifelse(insideout == "inside", "within cluster", "with entire background set") ) %>%
    mutate(insideout = factor(insideout, levels = c("within cluster", "with entire background set")))

ggplot(res.rbind, aes(x= as.factor(cluster1),y = value, color = insideout)) + geom_boxplot() + facet_wrap(~block) + theme_bw() + xlab("Cluster ID") + ylab("Proportionality distance") + labs(color = "Proportionality distance") + ggtitle("Proportionality distance per omic dataset") + ylim(-4, max(res.rbind$value))

```

We used a Wilcoxon test to compare the median within the cluster and outside the cluster.

```{r, echo=F, warning=F, message=F}
## all
all.bioR <- cbind(OTU.bioR, metabo.bioR)
phs.all <- propr::propr(all.bioR, metric = 'phs')@matrix

res.phs.all <- phs.all %>% as.data.frame %>% rownames_to_column("feature1") %>%
    gather(feature2, value, -feature1) %>%
    filter(feature1 %in% cluster.bioR$molecule) %>%
    filter(feature2 %in% cluster.bioR$molecule) %>%
    left_join(cluster.bioR  %>% dplyr::select(-block), 
              by = c("feature1" = "molecule")) %>%
    rename(cluster1 = cluster) %>% 
    left_join(cluster.bioR %>% dplyr::select(-block), 
              by = c("feature2" = "molecule")) %>%
    rename(cluster2 = cluster) %>%
    mutate(insideout = ifelse(cluster1 == cluster2, "inside", "outside")) %>%
        mutate(insideout = ifelse(insideout == "inside", "within cluster", "with entire background set") ) %>%
        mutate(insideout = factor(insideout, levels = c("within cluster", "with entire background set")))

ggplot(res.phs.all, aes(x= as.factor(cluster1),y = value, color = insideout)) + geom_boxplot(outlier.size = 0.5) + theme_bw() + xlab("Cluster ID") + ylab("Proportionality distance") + labs(color = "Proportionality distance") + ggtitle("Proportionality distance  (all omics)")

return_stat_median_only <- function(res.phs.X){
    i = 1
    res.pval <- matrix(ncol = 4, nrow = 4) %>% as.data.frame() %>%
    set_names("cluster", "median inside", "median outside", "Wilcoxon test Pval" )
    for(clu in unique(res.phs.X$cluster1)){
        inside <- res.phs.X %>% filter(cluster1 == clu) %>% filter(cluster2==clu) %>% pull(value)
        outside <- res.phs.X %>% filter(cluster1 == clu) %>% filter(cluster2!=clu) %>% pull(value)
        
        ttest.pval <- t.test(inside, outside)$p.value
        utest.pval <- wilcox.test(inside, outside)$p.value
        
        res.pval[i,] <- c(clu, round(median(inside), digits = 2),
                          round(median(outside), digits = 2), utest.pval)
        i = i+1
    }
    return(res.pval)
}
return_stat_median_only(res.phs.all) %>% knitr::kable()
```



# block sPLS

The three datasets (OTUs, metabolites and performances) are analysed together.
We used a block sparse PLS analysis.

```{r, echo=FALSE, warning=FALSE}
Liste=list(OTU=data.OTUS, metabolites=data.metabolites)
Y=cbind(perf.data, liq.data)
KeepX=list(OTU=c(20, 12),metabolites=c(8,8) ) # these values can be modified
KeepY=c(3, 1) # these values can be modified 
res.block = block.spls(X = Liste, Y=Y, indY = 3, ncomp = 2, keepX = KeepX, keepY = KeepY, mode = "regression")
```

The following graphs have been previously described.

```{r, echo=FALSE, warning=FALSE}

plotIndiv(res.block) #individuals plot
plotArrow(res.block)
plotVar(res.block,legend = TRUE, style ='graphics') #variables plot
```

```{r, echo=FALSE, warning=FALSE}
comp.block = compo.block(res.block)### function to extract the name of the selected variables in each block and get the name of the component and sign of the loading on that component 
var.selected.OTU = comp.block$var.selected.OTU
var.selected.metabolites = comp.block$var.selected.metabolites
var.selected.perf = comp.block$var.selected.perf

sign.OTU = comp.block$sign.OTU
sign.metabolites = comp.block$sign.metabolites
sign.perf = comp.block$sign.perf
```

```{r, echo=FALSE, warning=FALSE, eval=TRUE}
## modelled profiles of the selected variables are gathered in table for graphical purpose


X11=data.OTUS
melt.data1 = melt(t(rbind(
  cbind(scale(X11[, names(var.selected.OTU)])))))
melt.data1$comp= as.factor(rep(c(var.selected.OTU), nlevels(as.factor(c(10,13,16,19,22,25,28,31,34,37,40,43,46,50,53,57)))))
# filling up the melted values by repeating comp and sign 
melt.data1$sign= as.factor(rep(c(sign.OTU), nlevels(as.factor(c(10,13,16,19,22,25,28,31,34,37,40,43,46,50,53,57)))))
melt.data1$data = rep(c(rep('OTUs', length(var.selected.OTU))), nlevels(as.factor(c(10,13,16,19,22,25,28,31,34,37,40,43,46,50,53,57))))
# subset only on a particular component and represent pos / neg shapes 
m = length(melt.data1$data)/nlevels(as.factor(c(10,13,16,19,22,25,28,31,34,37,40,43,46,50,53,57)))
melt.data1$day = rep(c(10,13,16,19,22,25,28,31,34,37,40,43,46,50,53,57),each=m)

#print(ggplot(subset(melt.data1), aes(x=day, y=value, group = Var1, color = data)) +  geom_line() +facet_grid(sign~comp,scales="free") + ggtitle('sPLS') + xlab('time'))


X21=data.metabolites
melt.data2 = melt(t(rbind(
  cbind(scale(X21[, names(var.selected.metabolites)])))))
melt.data2$comp= as.factor(rep(c(var.selected.metabolites), nlevels(as.factor(c(10,13,16,19,22,25,28,31,34,37,40,43,46,50,53,57)))))
# filling up the melted values by repeating comp and sign 
melt.data2$sign= as.factor(rep(c(sign.metabolites), nlevels(as.factor(c(10,13,16,19,22,25,28,31,34,37,40,43,46,50,53,57)))))
melt.data2$data = rep(c(rep('Metabolites', length(var.selected.metabolites))), nlevels(as.factor(c(10,13,16,19,22,25,28,31,34,37,40,43,46,50,53,57))))
# subset only on a particular component and represent pos / neg shapes 
m = length(melt.data2$data)/nlevels(as.factor(c(10,13,16,19,22,25,28,31,34,37,40,43,46,50,53,57)))
melt.data2$day = rep(c(10,13,16,19,22,25,28,31,34,37,40,43,46,50,53,57),each=m)

#print(ggplot(subset(melt.data2), aes(x=day, y=value, group = Var1, color = data)) +  geom_line() +facet_grid(sign~comp,scales="free") + ggtitle('sPLS') + xlab('time'))


X31=cbind(perf.data, liq.data)
melt.data3 = melt(t(rbind(
  cbind(scale(X31[, names(var.selected.perf)])))))
melt.data3$comp= as.factor(rep(c(var.selected.perf), nlevels(as.factor(c(10,13,16,19,22,25,28,31,34,37,40,43,46,50,53,57)))))
# filling up the melted values by repeating comp and sign 
melt.data3$sign= as.factor(rep(c(sign.perf), nlevels(as.factor(c(10,13,16,19,22,25,28,31,34,37,40,43,46,50,53,57)))))
melt.data3$data = rep(c(rep('Performances', length(var.selected.perf))), nlevels(as.factor(c(10,13,16,19,22,25,28,31,34,37,40,43,46,50,53,57))))
# subset only on a particular component and represent pos / neg shapes 
m = length(melt.data3$data)/nlevels(as.factor(c(10,13,16,19,22,25,28,31,34,37,40,43,46,50,53,57)))
melt.data3$day = rep(c(10,13,16,19,22,25,28,31,34,37,40,43,46,50,53,57),each=m)

#print(ggplot(subset(melt.data3), aes(x=day, y=value, group = Var1, color = data)) +  geom_line() +facet_grid(sign~comp,scales="free") + ggtitle('sPLS') + xlab('time'))
```


```{r, echo=FALSE}

melt.rgcca=rbind(melt.data1, melt.data2, melt.data3)


color=c("tomato","slateblue", "chartreuse")
names(color)=c("OTUs","Metabolites", "Performances")

#png("blockspls_bioreactor.png", width = 3000, height = 1800, res = 400,pointsize=12)

##plot of the different scale profiles selected

print(ggplot(melt.rgcca, aes(x=Var2, y=value, group = interaction(Var1,data), color = data)) +scale_color_manual(values=color)+ geom_line(lwd=0.2) +   facet_grid(sign~comp, labeller=labeller(sign = c(pos="Positive",neg="Negative"), comp=c('1'="Component 1", '2'="Component 2", '3'='Component 3')))+ ggtitle('                         block sPLS, correlated data across time') + xlab('Number of days')+ scale_y_continuous(limits = c(-2, 2)) +ylab('Scaled value'))

#dev.off()


```
```{r, echo = F}
## variables selected with block spls
selected.variables=levels(melt.rgcca$Var1)
non.selected.OTUs=colnames(data.OTUS)[!colnames(data.OTUS)%in%selected.variables]
  non.selected.metabolites=colnames(data.metabolites)[!colnames(data.metabolites)%in%selected.variables]
```

## Proportionality analysis

```{r, echo=F, warning=F, message=F}
cluster.bioR <- cluster.block

OTU.bioR <- res.block$X$OTU %>% unscale %>% `+`(abs(min(.)))
phs.OTU <- propr::propr(OTU.bioR, metric = 'phs')@matrix
res.phs.OTU <- phs.OTU %>% as.data.frame %>% rownames_to_column("feature1") %>%
    gather(feature2, value, -feature1) %>%
    filter(feature1 %in% cluster.bioR$molecule) %>%
    filter(feature2 %in% cluster.bioR$molecule) %>%
    left_join(cluster.bioR %>% filter(block == "OTU") %>% dplyr::select(-block), 
              by = c("feature1" = "molecule")) %>%
    rename(cluster1 = cluster) %>%
    left_join(cluster.bioR %>% filter(block == "OTU"), 
              by = c("feature2" = "molecule")) %>%
    rename(cluster2 = cluster) %>%
    mutate(insideout = ifelse(cluster1 == cluster2, "inside", "outside"))

metabo.bioR <- res.block$X$metabo %>% unscale %>% `+`(abs(min(.)))
phs.metabo <- propr::propr(metabo.bioR, metric = 'phs')@matrix
res.phs.metabo <- phs.metabo %>% as.data.frame %>% rownames_to_column("feature1") %>%
    gather(feature2, value, -feature1) %>%
    filter(feature1 %in% cluster.bioR$molecule) %>%
    filter(feature2 %in% cluster.bioR$molecule) %>%
    left_join(cluster.bioR %>% filter(block == "Metabolites") %>% dplyr::select(-block), 
              by = c("feature1" = "molecule")) %>%
    rename(cluster1 = cluster) %>%
    left_join(cluster.bioR %>% filter(block == "Metabolites"), 
              by = c("feature2" = "molecule")) %>%
    rename(cluster2 = cluster) %>%
    mutate(insideout = ifelse(cluster1 == cluster2, "inside", "outside"))

perf.bioR <- res.block$X$Y %>% unscale %>% `+`(abs(min(.)))
phs.perf <- propr::propr(perf.bioR, metric = 'phs')@matrix
res.phs.perf <- phs.perf %>% as.data.frame %>% rownames_to_column("feature1") %>%
    gather(feature2, value, -feature1) %>%
    filter(feature1 %in% cluster.bioR$molecule) %>%
    filter(feature2 %in% cluster.bioR$molecule) %>%
    left_join(cluster.bioR %>% filter(block == "Y") %>% dplyr::select(-block), 
              by = c("feature1" = "molecule")) %>%
    rename(cluster1 = cluster) %>%
    left_join(cluster.bioR %>% filter(block == "Y"), 
              by = c("feature2" = "molecule")) %>%
    rename(cluster2 = cluster) %>%
    mutate(insideout = ifelse(cluster1 == cluster2, "inside", "outside"))

res.rbind <- rbind(res.phs.OTU, res.phs.metabo, res.phs.perf) %>%
    mutate(insideout = ifelse(insideout == "inside", "within cluster", "with entire background set") ) %>%
    mutate(insideout = factor(insideout, levels = c("within cluster", "with entire background set"))) %>%
    mutate(block = factor(block, levels = c('OTU', 'Metabolites', 'Y')))

ggplot(res.rbind, aes(x= as.factor(cluster1),y = value, color = insideout)) + geom_boxplot() + facet_wrap(~block) + theme_bw() + xlab("Cluster ID") + ylab("Proportionality distance") + labs(color = "Proportionality distance") + ggtitle("Proportionality distance per omic dataset") + ylim(-4, max(res.rbind$value))

## all
all.bioR <- cbind(OTU.bioR, metabo.bioR, perf.bioR)
phs.all <- propr::propr(all.bioR, metric = 'phs')@matrix

res.phs.all <- phs.all %>% as.data.frame %>% rownames_to_column("feature1") %>%
    gather(feature2, value, -feature1) %>%
    filter(feature1 %in% cluster.bioR$molecule) %>%
    filter(feature2 %in% cluster.bioR$molecule) %>%
    left_join(cluster.bioR  %>% dplyr::select(-block), 
              by = c("feature1" = "molecule")) %>%
    rename(cluster1 = cluster) %>% 
    left_join(cluster.bioR %>% dplyr::select(-block), 
              by = c("feature2" = "molecule")) %>%
    rename(cluster2 = cluster) %>%
    mutate(insideout = ifelse(cluster1 == cluster2, "inside", "outside")) %>%
        mutate(insideout = ifelse(insideout == "inside", "within cluster", "with entire background set") ) %>%
        mutate(insideout = factor(insideout, levels = c("within cluster", "with entire background set")))

ggplot(res.phs.all, aes(x= as.factor(cluster1),y = value, color = insideout)) + geom_boxplot(outlier.size = 0.5) + theme_bw() + xlab("Cluster ID") + ylab("Proportionality distance") + labs(color = "Proportionality distance") + ggtitle("Proportionality distance  (all omics)")

return_stat_median_only(res.phs.all) %>% knitr::kable()
```

